---
title: "Capstone Project Report"
author: "Emma Chandler"
date: "2025-05-30"
output: html_document
---

# Project Goal

The goal of this project was to map sequences from e. coli samples to the e. coli Genome and analyze the number of variant sites that differ between the samples and genomes.

## Link to GitHub Repository with scripts and summary statistic data
https://github.com/echandle2228/BINF8960_Capstone

## Analysis of summary statistics from Genomic pipeline

Load Data and Packages
```{r}
# Load Packages
library(tidyverse)

# Load data
stats <- read.csv("results/summary_stats.csv", header = TRUE)

```

Make Plots using summary statistics data
```{r}
# Plot showing read counts
stats %>% 
  filter(read_type != "variants") %>%
  ggplot() +
  aes(x = read_type, y = count, 
      group = sample,
      color = sample) +
  geom_line() +
  theme_classic() +
  labs(x = "", y = "Read Counts", color = "Sample") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0))

# Plot a bar chart of variants
stats %>% 
  filter(read_type == "variants") %>%
  ggplot() +
  aes(x = sample, y = count) +
  geom_col() +
  theme_classic() +
  labs(x = "Sample", y = "Variants") 


```

